Tail associations in ecological variables and their impact on extinction risk

### 2020-01-25 Response.txt

General comments
1.      The title seems a bit cryptic to readers who are not a statistician. I would suggest that title be modified to clarify it is about the tails of the distributions of ecological variables.

>>>Thank you for the suggestion. We changed the title as suggested.

2.      A conceptual schematic diagram early in the introduction will greatly help readers understand the main idea quickly and smoothly follow the text. In the current form, it is relatively hard to follow. Figure 1 does help, but a figure showing the idea in a more concrete way using tangible examples would make it much easier for readers to absorb the idea early on in the paper. It also would be helpful to see an example from finance or other fields that are somewhat analogous to ecological situations.

>>>Thank you for this suggestion, which spurred a lot of thought about how to best communicate the main ideas of the manuscript in the intial parts of the manuscript. The referee's suggestion led to two new figures.

@@

3.      The amount of analyses for metapopulations and age-structured populations seems quite unbalanced. Why is this so? Obviously, the paper would greatly increase its value by including more analyses for stage-structured populations, but then the paper may become too long to fit in the required length for Ecosphere. Would it be more beneficial to focus only on metapopulations in this paper and expand on the stage-structured population part in a separate paper to more deeply examine Tuljapurkar’s approximation?

>>>Thank you for asking these questions. First, this is an initial attempt to understand the impact of asymmetric tail associations on extinction risk, and as such we intentionally used a simple approach for a rapid assessment of whether the basic ideas have merit. Based on our results, we believe this could turn into a broad topic, which many researchers should investigate. Our intention was to provide an initial exploration of the mechanisms rapidly, and in so doing, hopefully to open what could be a practically important topic (for conservation) to consideration by other researchers, as quickly and broadly as possible. With regard to extensive mathematical analyses (e.g. using Tuljapurkar's approaximation) of stochastic matrix models, which the referee suggests, we are journeymen only, not true experts. Mathematical analysis of stochastic matrix models is a deep field. So we believe carving our matrix model results out of the current paper and then undertaking a more systematic explortion ourselves, for possible later publication, does not get these potentially practically important ideas out there as fast and risks the possibility that we will not be able to do the analyses as well as some of our readers. So to us it is important to get these results out there now.
  Second, for stochastic matrix models, our initial explorations suggested that probably the effects of tail association in that context are minor, so we went no further. In contrast, for metapopulation models, our initial explorations (initially our explorations were with the Ricker model only) showed a likely strong effect, so we went further with additional models. So what the referee perceives as imbalance is actually just a rational attempt to explore promising ideas further and to simply identify other ideas as less promising, to help future researchers choose what to work on.  
  Finally, we do not see as much of an imbalance as the referee suggests. To us, the two components are treated with slightly different throroughness, but not markedly so, and for good reasons (see above). Our matrix modelling results are certainly enough to make the points we wanted to make, i.e., an initial assessment that the influence of tail associations on extinction risk in this context seems likely to be minor.

@@

4.      The order of appearance of the metapopulation problem and the age-structured population problem keeps switching throughout the paper. For readability, the order should be consistent.

>>>Thanks for raising this point. We have edited to enforce consistency of this ordering throughout the manuscript.

5.      Whether variables have tail associations can be altered by transformation; e.g., variables may have no tail association in one way but show any of the tail associations discussed in this paper in another. How should researchers who want to use the approached introduced in this paper deal with this? Should they look for a transformation that can give one of such tail associations as long as it is ecologically meaningful in a given system? I suppose so, but this is not mentioned in the paper.

>>>The referee is mistaken here, transformation does not alter tail associations as we have defined them based on data ranks. This is a standard way of defining tail associations and copula approaches to the study of dependencies, justified by the Ghosh et al paper (which, by the way, has now been formally accepted at Advances in Ecological Research, and is in press) and by Genest & Favre, 2007, J Hydol Eng, "Everything you always wanted to know about copula modelling but were afraid to ask". We have clarified and amended some text, in the Discussion, which helps explain these facts.

@@

Also, the interpretation that certain variables are “bad” or “good” depends on how it is formulated or transformed. It could be that, if a “bad” variable is associated with a “good” variable in their right tails, for example, their effects cancel each other and leave no overall effect, which would be also an interesting result. This may be related to the discussion about negative associations between Line 411 and 423, but it does not clearly come across. Maybe these are discussed in the Ghosh et al. paper in BioRxiv, but it is not yet published in a peer reviewed journal. These aspects should be worth mentioning in this paper for clarification of the method.

>>>The Ghosh et al paper is now formally accepted. The version on the BioRxiv is essentially the same as the final, accepted version. In spite of thinking for a quite a long time about what the referee may have meant by the above comment, we confess we are not sure, so we do not know quite what to do about this comment. In our model formulations, variables themselves are never "good" or "bad" but are instead assumed to fluctuate, taking values in some years which are "good" for the population (by which we mean they promote increased reproduction or increased survival in that location, in that year), and values in other years which are "bad" (in the opposite sense). Aside from an apparent desire to have us make some additional point (I'm afraid we cannot understand what point), the referee does not appear to be criticizing anything we actually do say. So we suggest proceeding with this part of the manuscript as it is. As we said, the Ghosh et al paper is now accepted, and is already available online on the BioRxiv, and many important  points which seem likely to be related to the referee's point are made there. So readers looking to base some of their own research on the points made in the current manuscript can also use the Ghosh et al paper as an additional background resource before getting started.

@@

•       As I mentioned above, Q1 and Q2 appear in an opposite order from what I expected from the introduction so far.

>>>We have fixed this ordering issue throughout the manuscript, see our response to point 4 above.

•       Line 299. Because simulation studies can increase the number of replicates relatively easily and usually use a large number of replicates, statistical significance does not mean much in most of cases, including this one.

>>>Thank you, we take the referee's point, here, and have eliminated the statement about significance.

### 2020-01-27 Response.txt

Dear editors,

Thank you for the attention and expertise you used in assessing, for Ecosphere, our paper "Tail associations and their impact on extinction risk" (this was the old title, we will now be changing it in response to a referee request - see below). Thank you also for the generally very positive reviews. We have responded to all referee comments through changes to the manuscript, and below. We find the paper is improved and we hope you now find it ready for formal acceptance to Ecosphere. Our responses to comments are preceded by ">>>" to make them easy to find. Thank you again for an enjoyable and helpful review process.

@@

General comments
1.      The title seems a bit cryptic to readers who are not a statistician. I would suggest that title be modified to clarify it is about the tails of the distributions of ecological variables.

>>>Thank you for the suggestion. We changed the title as suggested.

2.      A conceptual schematic diagram early in the introduction will greatly help readers understand the main idea quickly and smoothly follow the text. In the current form, it is relatively hard to follow. Figure 1 does help, but a figure showing the idea in a more concrete way using tangible examples would make it much easier for readers to absorb the idea early on in the paper. It also would be helpful to see an example from finance or other fields that are somewhat analogous to ecological situations.

>>>Thank you for this suggestion, which spurred a lot of thought about how to best communicate the main ideas of the manuscript in the initial parts of the manuscript. The referee's suggestion led to two new figures.

@@

3.      The amount of analyses for metapopulations and age-structured populations seems quite unbalanced. Why is this so? Obviously, the paper would greatly increase its value by including more analyses for stage-structured populations, but then the paper may become too long to fit in the required length for Ecosphere. Would it be more beneficial to focus only on metapopulations in this paper and expand on the stage-structured population part in a separate paper to more deeply examine Tuljapurkar’s approximation?

>>>Thank you for asking these questions. First, this is an initial attempt to understand the impact of asymmetric tail associations on extinction risk, and as such we intentionally used a simple approach for a rapid assessment of whether the basic ideas have merit. Based on our results, we believe this could turn into a broad topic, which many researchers should investigate. Our intention was to provide an initial exploration of the mechanisms rapidly, and in so doing, hopefully to open what could be a practically important topic (for conservation) to consideration by other researchers, as quickly and broadly as possible. With regard to extensive mathematical analyses (e.g. using Tuljapurkar's approaximation) of stochastic matrix models, which the referee suggests, we are journeymen only, not true experts. Mathematical analysis of stochastic matrix models is a deep field. So we believe carving our matrix model results out of the current paper and then undertaking a more systematic exploration ourselves, for possible later publication, does not get these potentially practically important ideas out there as fast and risks the possibility that we will not be able to do the analyses as well as some of our readers. So to us it is important to get these results out there now.
  Second, for stochastic matrix models, our initial explorations suggested that probably the effects of tail association in that context are minor, so we went no further. In contrast, for metapopulation models, our initial explorations (initially our explorations were with the Ricker model only) showed a likely strong effect, so we went further with additional models. So what the referee perceives as imbalance is actually just a rational attempt to explore promising ideas further and to simply identify other ideas as less promising, to help future researchers choose what to work on.  
  Finally, we do not see as much of an imbalance as the referee suggests. To us, the two components are treated with slightly different thoroughness, but not markedly so, and for good reasons (see above). Our matrix modelling results are certainly enough to make the points we wanted to make, i.e., an initial assessment that the influence of tail associations on extinction risk in this context seems likely to be minor.

### 2020-01-31 MainText.Rmd
cleaned pto fig legend
table cleaned
Popovic Ref updated

<!--future ideas start here-->
Our exploration for stage-structured models suggested limited influence of 
tail associations between vital rates on extinction risk, but some
additional lines of inquiry might be worth exploring as part of an effort to
understand how general our results are.
First, we considered a model with positive association between 
egg and adult vital rates,
corresponding to similar impacts of an environmental variable on 
eggs and adults. Negative associations between vital rates
can occur as well, and the impacts of tail associations should be considered
in that context as well. Negative associations relate to
life history trade-offs, e.g. between survival and reproduction 
(e.g, @Lande2003, p. 59). Asymmetric tail associations may occur
in such a context if, for instance, there is a three way trade-off
among vital rates $r_1$, $r_2$, and $r_3$. Values of $r_1$ near
an upper limit may be strongly associated with $r_2$, whereas lower 
values of $r_1$ may not be strongly
associated with $r_2$ because $r_1$ trades off against both 
$r_2$ and $r_3$. It is possible that tail associations in negatively 
related vital rates produce different outcomes for extinction risk
from what we observed by studying positively associated vital rates.

@@

Rank-based methods are linked to a well-developed suite of statistical
approaches related to the *copula* concept of statistics. Copulas and related ideas
can be used to formally separate the information in the joint 
distribution of $X_i$ and $X_j$ into the information that pertains
to the marginal distributions of these variables, and the information
that pertains solely to their association. 
@Ghosh_copula argued generally for the potential of copulas 
to improve our understanding of ecology 
(see also @Popovic2018; @Valpine2014; @Anderson2018).
Rank-based methods such as the Spearman and partial Spearman correlations
we used are closely related to copulas, and @Ghosh_copula,
@Genest2007 and others have recommended copula and rank approaches when the 
goal is to understand associations between variables. We recognize that the Pearson correlation
of $X_i(t)$ and $X_j(t)$, and changes through time in that correlation, 
will still be useful for many applications. But to understand potential 
changes in the association 
of $X_i(t)$ and $X_j(t)$, unconfounded by changes in their marginal distributions,
it is necessary to use rank-based or copula methods [@Ghosh_copula;
@Genest2007].

@@

<!--Summary tables for 6 nonlinear models used-->
```{r read_tab_model_summary,echo=F, results='asis',message=F}
library(kableExtra)
library(dplyr)
dt<-readRDS("./Results/tab_model_summary.RDS")
knitr::kable(dt, escape=F,
             format="latex",digits=3, align="c",table.env='table*',
             caption = "Summary of the six density-dependent stochastic metapopulation models we used. The population growth rate in habitat patch $i$ at time $t$, $\\lambda_i(t)$ (second column), is $P_i(t+1)/P_i(t)$, and defines how the model works within each patch. Here $P_i(t)$ is the population in patch $i$ at time $t$. The growth rate depends on model parameters, $P_i(t)$, and the environment $\\varepsilon_i(t)$ in patch $i$ at time $t$. $\\varepsilon_i(t)$ was normally distributed with mean $0$ and standard deviation $\\sigma$, and values were generated independently across times. $P_e$ is the equilibrium of the deterministic, one-patch version of the model. $r_{min}$ is the smallest $r$, given fixed values of the other parameters, for which $P_e$ was non-negative. $r_c$ is the value of $r$ for which a transition occurs from under- to overcompensatory dynamics (see main text). $r_{bf}$ is the value of $r$ for which a transition occurs from a stable to an unstable equilbrium (a bifurcation; see main text). For extinction-risk simulations, we used $r$ such that $r_{min}<r<r_{bf}$. Dispersal between patches was also implemented (see main text). Formulas are given for $P_e$ but specific values which pertain for the listed parameters are given for $r_{min}$, $r_c$ and $r_{bf}$. See Appendix \\ref{SM-model_stability} for the formulas that gave these values.\\label{tab_model_summary}",
             booktabs=T, linesep = "\\addlinespace"
             )%>%column_spec(c(1,3,5,6,7),width="1.4 cm")%>%kable_styling(latex_options = "basic",position = "center")%>%column_spec(2, width = "3.8 cm")
```

>>>>

<!--future ideas start here-->
Our exploration for stage-structured models suggested limited influence of 
tail associations between vital rates on extinction risk, but some
additional lines of inquiry might be worth exploring as part of an effort to
understand how general our results are.
First, we considered a model with positive association between 
egg and adult vital rates,
corresponding to similar impacts of an environmental variable on 
eggs and adults. Negative associations between vital rates
can occur as well, and the impacts of tail associations should be considered
in that context as well. Negative associations relate to
life history trade-offs, e.g. between survival and reproduction 
(e.g., @Lande2003, p. 59). Asymmetric tail associations may occur
in such a context if, for instance, there is a three way trade-off
among vital rates $r_1$, $r_2$, and $r_3$. Values of $r_1$ near
an upper limit may be strongly associated with $r_2$, whereas lower 
values of $r_1$ may not be strongly
associated with $r_2$ because $r_1$ trades off against both 
$r_2$ and $r_3$. It is possible that tail associations in negatively 
related vital rates produce different outcomes for extinction risk
from what we observed by studying positively associated vital rates.

@@

Rank-based methods are linked to a well-developed suite of statistical
approaches related to the *copula* concept of statistics. Copulas and related ideas
can be used to formally separate the information in the joint 
distribution of $X_i$ and $X_j$ into the information that pertains
to the marginal distributions of these variables, and the information
that pertains solely to their association. 
@Ghosh_copula argued generally for the potential of copulas 
to improve our understanding of ecology 
(see also @Popovic2019; @Valpine2014; @Anderson2018).
Rank-based methods such as the Spearman and partial Spearman correlations
we used are closely related to copulas, and @Ghosh_copula,
@Genest2007 and others have recommended copula and rank approaches when the 
goal is to understand associations between variables. We recognize that the Pearson correlation
of $X_i(t)$ and $X_j(t)$, and changes through time in that correlation, 
will still be useful for many applications. But to understand potential 
changes in the association 
of $X_i(t)$ and $X_j(t)$, unconfounded by changes in their marginal distributions,
it is necessary to use rank-based or copula methods [@Ghosh_copula;
@Genest2007].

@@

<!--Summary tables for 6 nonlinear models used-->
```{r read_tab_model_summary,echo=F, results='asis',message=F}
library(kableExtra)
library(dplyr)
dt<-readRDS("./Results/tab_model_summary.RDS")
knitr::kable(dt, escape=F,
             format="latex",digits=3, align="c",table.env='table*',
             caption = "Summary of the six density-dependent stochastic metapopulation models we used. The population growth rate in habitat patch $i$ at time $t$, $\\lambda_i(t)$ (second column), is $P_i(t+1)/P_i(t)$, and defines how the model works within each patch. Here $P_i(t)$ is the population in patch $i$ at time $t$. The growth rate depends on model parameters, $P_i(t)$, and the environment $\\varepsilon_i(t)$ in patch $i$ at time $t$. $\\varepsilon_i(t)$ was normally distributed with mean $0$ and standard deviation $\\sigma$, and values were generated independently across times. $P_e$ is the equilibrium of the deterministic, one-patch version of the model. $r_{min}$ is the smallest $r$, given fixed values of the other parameters, for which $P_e$ was non-negative. $r_c$ is the value of $r$ for which a transition occurs from under- to overcompensatory dynamics (see main text). $r_{bf}$ is the value of $r$ for which a transition occurs from a stable to an unstable equilibrium (a bifurcation; see main text). For extinction-risk simulations, we used $r$ such that $r_{min}<r<r_{bf}$. Dispersal between patches was also implemented (see main text). Formulas are given for $P_e$ but specific values which pertain for the listed parameters are given for $r_{min}$, $r_c$ and $r_{bf}$. See Appendix \\ref{SM-model_stability} for the formulas that gave these values.\\label{tab_model_summary}",
             booktabs=T, linesep = "\\addlinespace"
             )%>%column_spec(c(1,3,5,6,7),width="1.4 cm")%>%kable_styling(latex_options = "basic",position = "center")%>%column_spec(2, width = "3.8 cm")
```
